---
title: "EOV Coverage Across Three ONMS Products"
format:
  html:
    toc: true
    embed-resources: true
    code-fold: true
    code-tools: true
execute:
  echo: true
  message: true
  warning: true
  error: true
editor_options:
  chunk_output_type: console
---

This summary interrogates three sister repos for the latest variable coverage
mapped onto proposal Table 1 Essential Ocean Variables (EOVs):

- **Shiny app** — this repo (`climate-dashboard-app`): `app/global.R` `d_vars` +
  `meta/*.yml`
- **flexdashboard** — `../climate-dashboard`: `data/datasets.csv`
- **eco-indicators** — `../eco-indicators`: `data/copernicus_variables_trim.csv` +
  `data/erddap.csv`

> Render this on the host where `climate-dashboard-app` is deployed to capture
> the latest from live data connections. Sibling repos (`../eco-indicators`,
> `../climate-dashboard`) must be checked out alongside at their latest commits.

```{r}
#| label: setup
#| warning: false
librarian::shelf(
  dplyr,
  DT,
  fs,
  ggplot2,
  glue,
  here,
  purrr,
  readr,
  stringr,
  tibble,
  tidyr,
  yaml,
  quiet = T
)
options(readr.show_col_types = F)

dir_app <- here()
dir_eco <- here("../eco-indicators")
dir_dash <- here("../climate-dashboard")
out_csv <- here("data/summary_eov_coverage.csv")
dir.create(dirname(out_csv), showWarnings = F, recursive = T)
```

## Canonical EOV list

```{r}
#| label: eov_levels
eov_desc <- tribble(
  ~eov                        , ~description                                                                                                                                                                       ,
  "SST"                       , "Sea surface temperature — near-surface ocean temperature (°C)."                                                                                                                   ,
  "Chl-a"                     , "Chlorophyll-a concentration — proxy for phytoplankton biomass (mg/m³)."                                                                                                           ,
  "SSS"                       , "Sea surface salinity — near-surface ocean salinity (g/kg or psu)."                                                                                                                ,
  "MLD"                       , "Mixed layer depth — thickness of the well-mixed upper ocean (m)."                                                                                                                 ,
  "Bottom-T"                  , "Bottom temperature — sea water potential temperature at the sea floor (°C)."                                                                                                      ,
  "Oxygen"                    , "Dissolved oxygen — mole concentration of molecular O₂ in sea water (mmol/m³)."                                                                                                    ,
  "pH / DIC"                  , "Ocean acidification indicators — pH, dissolved inorganic carbon, and surface pCO₂."                                                                                               ,
  "NPP"                       , "Net primary production — rate of carbon fixation by phytoplankton (mg C/m²/day or mmol/m²/day)."                                                                                  ,
  "Nutrients (NO3/PO4/Si/Fe)" , "Macro-/micro-nutrients — nitrate, phosphate, silicate, iron (mmol/m³)."                                                                                                           ,
  "Phytoplankton (PFTs)"      , "Phytoplankton functional types — size classes (micro/nano/pico) and taxa (diatoms, dinoflagellates, haptophytes, greens, prochlorophytes, prokaryotes) as chlorophyll or carbon." ,
  "Zooplankton"               , "Zooplankton biomass — mesozooplankton carbon concentration (mmol/m³)."                                                                                                            ,
  "Precipitation"             , "Precipitation rate — freshwater flux from IMERG monthly product (mm)."                                                                                                            ,
  #"SSH"                       , "Sea surface height — sea level above geoid (m)."
)

eov_levels <- eov_desc$eov
product_levels <- c("Shiny app", "flexdashboard", "eco-indicators")
```

## Spatial-resolution helper

```{r}
#| label: parse_spatial_res
parse_spatial_res <- function(dataset_id) {
  case_when(
    str_detect(dataset_id, "0\\.083deg") ~ "0.083° (~9.25 km)",
    str_detect(dataset_id, "0\\.25deg") ~ "0.25° (~28 km)",
    str_detect(dataset_id, "4km") ~ "4 km",
    str_detect(dataset_id, "IMERG") ~ "0.1° (~11 km)",
    str_detect(dataset_id, "jplMURSST") ~ "0.01° (~1 km)",
    str_detect(dataset_id, "erdMH1chla") ~ "0.042° (~4.6 km)",
    str_detect(dataset_id, "noaacrwsst") ~ "0.05° (~5 km)",
    str_detect(
      dataset_id,
      regex("cesm2", ignore_case = T)
    ) ~ "~1° (CESM2 LENS2)",
    TRUE ~ NA_character_
  )
}
```

## EOV mapping

```{r}
#| label: eov_map
eov_map <- tribble(
  ~variable_key   , ~eov                        ,
  # Copernicus Physics (shared by Shiny app & eco-indicators)
  "thetao"        , "SST"                       ,
  "so"            , "SSS"                       ,
  "mlotst"        , "MLD"                       ,
  "bottomT"       , "Bottom-T"                  ,
  #"zos"           , "SSH"                       ,
  # Copernicus BGC (Shiny app)
  "o2"            , "Oxygen"                    ,
  "nppv"          , "NPP"                       ,
  "dissic"        , "pH / DIC"                  ,
  "ph"            , "pH / DIC"                  ,
  "spco2"         , "pH / DIC"                  ,
  "no3"           , "Nutrients (NO3/PO4/Si/Fe)" ,
  "po4"           , "Nutrients (NO3/PO4/Si/Fe)" ,
  "si"            , "Nutrients (NO3/PO4/Si/Fe)" ,
  "fe"            , "Nutrients (NO3/PO4/Si/Fe)" ,
  "chl"           , "Chl-a"                     ,
  "phyc"          , "Phytoplankton (PFTs)"      ,
  "zooc"          , "Zooplankton"               ,
  # Copernicus GlobColour (eco-indicators)
  "CHL"           , "Chl-a"                     ,
  "PP"            , "NPP"                       ,
  "DIATO"         , "Phytoplankton (PFTs)"      ,
  "DINO"          , "Phytoplankton (PFTs)"      ,
  "GREEN"         , "Phytoplankton (PFTs)"      ,
  "HAPTO"         , "Phytoplankton (PFTs)"      ,
  "MICRO"         , "Phytoplankton (PFTs)"      ,
  "NANO"          , "Phytoplankton (PFTs)"      ,
  "PICO"          , "Phytoplankton (PFTs)"      ,
  "PROCHLO"       , "Phytoplankton (PFTs)"      ,
  "PROKAR"        , "Phytoplankton (PFTs)"      ,
  # ERDDAP (Shiny app & flexdashboard & eco-indicators)
  "erddap_sst"    , "SST"                       ,
  "erddap_sss"    , "SSS"                       ,
  "erddap_precip" , "Precipitation"             ,
  "precipitation" , "Precipitation"             ,
  "sst"           , "SST"                       ,
  "chlorophyll"   , "Chl-a"                     ,
  # flexdashboard CESM2 short names
  "SST"           , "SST"                       ,
  "Chl"           , "Chl-a"                     ,
  "PH"            , "pH / DIC"                  ,
  "HMXL_DR"       , "MLD"                       ,
  "SSS"           , "SSS"                       ,
  "NPP"           , "NPP"
)
```

## Read Shiny app (`climate-dashboard-app`)

Extract `d_vars` from `app/global.R` (parse the `tribble(...)` literal to avoid
running the Shiny/bslib init) and augment with per-dataset metadata from
`meta/*.yml`.

```{r}
#| label: read_app
extract_d_vars <- function(global_r) {
  lines <- readLines(global_r)
  i_start <- grep("^d_vars\\s*<-\\s*tribble", lines)[1]
  opens <- cumsum(str_count(lines[i_start:length(lines)], "\\("))
  closes <- cumsum(str_count(lines[i_start:length(lines)], "\\)"))
  i_end <- i_start + which(opens == closes & opens > 0)[1] - 1
  txt <- paste(lines[i_start:i_end], collapse = "\n")
  eval(parse(text = txt))
}

d_vars <- extract_d_vars(file.path(dir_app, "app/global.R"))

read_meta_yml <- function(yml_path) {
  y <- read_yaml(yml_path)
  key <- path_ext_remove(path_file(yml_path))
  dsid <- paste(
    map_chr(y$datasets %||% list(list(id = NA)), "id"),
    collapse = "; "
  )
  vars <- y$variables %||% list()
  if (length(vars) == 0) {
    return(tibble(
      meta_key = key,
      var_id = NA_character_,
      var_name = NA_character_,
      units = NA_character_,
      dataset_id = dsid,
      depth_min = y$depth$min %||% NA_real_,
      depth_max = y$depth$max %||% NA_real_
    ))
  }
  tibble(
    meta_key = key,
    var_id = map_chr(vars, \(v) v$id %||% NA_character_),
    var_name = map_chr(vars, \(v) v$name %||% NA_character_),
    units = map_chr(vars, \(v) v$units %||% NA_character_),
    dataset_id = dsid,
    depth_min = y$depth$min %||% NA_real_,
    depth_max = y$depth$max %||% NA_real_
  )
}

meta_yml <- dir_ls(file.path(dir_app, "meta"), glob = "*.yml") |>
  map_dfr(read_meta_yml)

# split d_vars$var into meta_key + var_id (var_id = NA for erddap_*)
vars_app <- d_vars |>
  mutate(
    meta_key = if_else(
      str_detect(var, "\\."),
      str_extract(var, "^[^.]+"),
      var
    ),
    var_id = if_else(
      str_detect(var, "\\."),
      str_extract(var, "(?<=\\.).+$"),
      NA_character_
    )
  ) |>
  left_join(meta_yml, by = c("meta_key", "var_id")) |>
  mutate(
    product = "Shiny app",
    variable_key = coalesce(var_id, var),
    dataset = coalesce(dataset_id, meta_key),
    spatial_res = parse_spatial_res(dataset_id),
    time_min = NA_character_,
    time_max = NA_character_
  ) |>
  select(
    product,
    variable = var,
    variable_key,
    label,
    dataset,
    spatial_res,
    time_min,
    time_max,
    units
  )
```

## Read flexdashboard (`climate-dashboard`)

```{r}
#| label: read_dash
d_dash <- read_csv(
  file.path(dir_dash, "data/datasets.csv"),
  col_types = cols(active = col_character(), .default = col_guess())
) |>
  filter(active == "T")

vars_dash <- d_dash |>
  mutate(
    product = "flexdashboard",
    variable = var,
    variable_key = var,
    label = coalesce(var_header, plot_label),
    dataset = case_when(
      type == "erddap" ~ str_extract(url, "(?<=griddap/)[^.]+"),
      type == "nc" ~ "CESM2 LENS2",
      TRUE ~ NA_character_
    ),
    spatial_res = parse_spatial_res(coalesce(dataset, url)),
    time_min = if_else(!is.na(yr_beg), as.character(yr_beg), NA_character_),
    time_max = if_else(!is.na(yr_end), as.character(yr_end), NA_character_),
    units = str_extract(plot_label, "\\(.*\\)") |>
      str_remove_all("[()]")
  ) |>
  select(
    product,
    variable,
    variable_key,
    label,
    dataset,
    spatial_res,
    time_min,
    time_max,
    units
  )
```

## Read eco-indicators

```{r}
#| label: read_eco
copernicus_csv <- file.path(dir_eco, "data/copernicus_variables_trim.csv")
erddap_csv <- file.path(dir_eco, "data/erddap.csv")

d_cop <- read_csv(copernicus_csv) |>
  filter(!str_detect(variable, "_uncertainty"), variable != "flags")

vars_eco_cop <- d_cop |>
  distinct(
    dataset_id,
    product_title,
    variable,
    standard_name,
    units,
    time_min,
    time_max
  ) |>
  mutate(
    product = "eco-indicators",
    variable_key = variable,
    label = standard_name,
    dataset = dataset_id,
    spatial_res = parse_spatial_res(dataset_id)
  ) |>
  select(
    product,
    variable,
    variable_key,
    label,
    dataset,
    spatial_res,
    time_min,
    time_max,
    units
  )

d_erd <- read_csv(erddap_csv)
vars_eco_erd <- d_erd |>
  distinct(dataset, var) |>
  mutate(
    product = "eco-indicators",
    variable = var,
    variable_key = var,
    label = "Precipitation",
    spatial_res = parse_spatial_res(dataset),
    time_min = format(min(as.Date(d_erd$time)), "%Y-%m-%d"),
    time_max = format(max(as.Date(d_erd$time)), "%Y-%m-%d"),
    units = "mm"
  ) |>
  select(
    product,
    variable,
    variable_key,
    label,
    dataset,
    spatial_res,
    time_min,
    time_max,
    units
  )

vars_eco <- bind_rows(vars_eco_cop, vars_eco_erd)
```

## Combine & assign EOVs

```{r}
#| label: combine
vars_all <- bind_rows(vars_app, vars_dash, vars_eco) |>
  left_join(eov_map, by = "variable_key") |>
  filter(!is.na(eov), eov %in% eov_levels) |>
  mutate(
    product = factor(product, levels = product_levels),
    eov = factor(eov, levels = eov_levels)
  )

write_csv(vars_all, out_csv)
```

## EOV descriptions

```{r}
#| label: tbl_eov_desc
datatable(
  eov_desc,
  rownames = FALSE,
  options = list(pageLength = 20, dom = "t"),
  colnames = c("EOV", "Description")
)
```

## Coverage figure

```{r}
#| label: fig_coverage
#| fig-width: 7
#| fig-height: 5.5
coverage <- expand_grid(
  product = factor(product_levels, levels = product_levels),
  eov = factor(eov_levels, levels = eov_levels)
) |>
  left_join(
    vars_all |> distinct(product, eov) |> mutate(status = "live"),
    by = c("product", "eov")
  ) |>
  mutate(status = coalesce(status, "gap"))

ggplot(coverage, aes(product, eov, fill = status)) +
  geom_tile(color = "white", linewidth = 0.6) +
  scale_fill_manual(values = c(live = "#1f77b4", gap = "#d9d9d9")) +
  scale_y_discrete(limits = rev(eov_levels)) +
  labs(
    title = "EOV coverage across three products",
    subtitle = "Rows = proposal Table 1 variables · blue = live · grey = gap",
    x = NULL,
    y = NULL,
    fill = NULL
  ) +
  theme_minimal(base_size = 12) +
  theme(
    panel.grid = element_blank(),
    legend.position = "bottom"
  )
```

## Detail table

One row per (EOV, product) with the underlying variable codes, source dataset,
spatial resolution, and timespan. Export: [`data/summary_eov_coverage.csv`](data/summary_eov_coverage.csv).

```{r}
#| label: tbl_detail
detail <- vars_all |>
  group_by(eov, product) |>
  summarize(
    variables = paste(unique(variable), collapse = ", "),
    dataset = paste(unique(dataset), collapse = "; "),
    spatial_res = paste(unique(na.omit(spatial_res)), collapse = "; "),
    timespan = paste(
      na.omit(unique(paste0(time_min, " → ", time_max))),
      collapse = "; "
    ),
    .groups = "drop"
  ) |>
  left_join(eov_desc, by = "eov") |>
  relocate(description, .after = eov) |>
  arrange(eov, product)

datatable(
  detail,
  rownames = FALSE,
  options = list(scrollX = TRUE, pageLength = 25)
)
```
